---
title: "Lecture 7"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# R functions
We can source any file of R code with the `source`() function

```{r}
source("http://tinyurl.com/rescale-R")
```


Let's make sure things are here
```{r}
ls()
```

Check our `rescale()` is working
```{r}
rescale(1:10)
```

```{r,eval=FALSE}
rescale( c(1:10, "string"))
```


Let's check if `rescale2()` does any better
```{r, eval=FALSE}
rescale2( c(1:10, "string"))
```

## FUnction for finding missing values in two datasets

Write a `both_na()` function to do this

```{r}
x <- c( 1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3, 4)

is.na(x)

which(is.na(x))
which(!is.na(x))
```

```{r}
# how many time it is true in x
sum( is.na(x))

# how many times it is true in both x and y
sum( is.na(x) + is.na(y))

# where is it `na` in both x and y
is.na(x) & is.na(y)

# how many times is it `na` in both x and y
sum( is.na(x) & is.na(y) )
```


Writing a working function from this snippet
```{r}
both_na <- function(x, y) {
 sum( is.na(x) & is.na(y) )
}
```

Test it

```{r}
both_na(x, y)
```

```{r,eval=FALSE}
x <- c(NA, NA, NA)
y1 <- c( 1, NA, NA)
y2 <- c( 1, NA, NA, NA)

both_na(x, y2)
```

Improving function with error message
```{r}
both_na2 <- function(x, y) {
  
   if(length(x) != length(y)) {
    stop("Input x and y should be the same length")
   }
  
 sum( is.na(x) & is.na(y) )
}
```

Test new function
```{r, eval=FALSE}
both_na2(x, y2)
```


Refine function by adding more info `both_na3()`
```{r}
x <- c( 1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3, 4)

both_na3(x, y)
```





## And a last function
```{r}
x <- df1$IDs
y <- df2$IDs

x
y
```

```{r}
# find similarities between x and y
intersect(x, y)
```


We can try the `intersect()` function and the `%in%` functions
```{r}
intersect(x, y)
x %in% y
```


```{r}
x[x %in% y]
y[ y %in% x ]

#put these together as columns of a matrix
cbind( x[ x %in% y ], y[ y %in% x ] )
```

```{r}
#cbind
cbind( c("Hello", "Help"), c("Please", "Help"))

#rbind
rbind( c("Hello", "Help"), c("Please", "Help"))
```

Writing it into a function
```{r}
gene_intersect <- function(x, y) {
 cbind( x[ x %in% y ], y[ y %in% x ] )
}
```


```{r}
gene_intersect(x, y)
```


Gene intersect for data frames
```{r}
gene_intersect2(df1, df2)
```

Make the output look better, into `gene_intersect3`
```{r}
gene.colname="IDs"
df1[,gene.colname]

gene_intersect3(df1, df2)
```



Make the function look better
```{r}
gene_intersect4(df1, df3)
```


Let's use the `merge()` function
```{r}
merge(df1, df2, by="IDs")
```















