---
title: "lecture14"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Import all data
```{r}
counts <- read.csv("airway_scaledcounts.csv", stringsAsFactors = FALSE)
metadata <-  read.csv("airway_metadata.csv", stringsAsFactors = FALSE)
```

Look at the data
```{r}
head(counts)
head(metadata)
```

```{r}
colnames(counts)[-1] == metadata$id
#does the column names equal the id's of meta data?
```

```{r}
control <- metadata[metadata[,"dex"]=="control",]
control
```

```{r}
control.mean <- rowSums(counts[,control$id])/4
#takes sums of rows and divide by number of rows to get means
```

```{r}
names(control.mean) <- counts$ensgene
#adds ensemble gene names back to better keep track of gene data
head(control.mean)
```

Follow the same procedure for treated samples
```{r}
#get only treated samples
treated <- metadata[metadata[,"dex"]=="treated",]
treated

#take sums of treated rows and divide by number of rows
treated.mean <- rowSums(counts[,treated$id])/nrow(treated)
#divide by nrow(treated) to make the code more robust

names(treated.mean) <- counts$ensgene

head(treated.mean)
```

Combine into new dataframe
```{r}
meancounts <- data.frame(control.mean, treated.mean)
head(meancounts)
colSums(meancounts)
```

Plot the data
```{r}
plot(meancounts, xlab = "log Control", ylab = "log Treated", 
     log = "xy")
#plot as log to see all points
```

Calculate log2foldchange
```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```
See some weird results. "NaN" is not a number and -Inf results
This is caused by genes with 0 expression

```{r}
x <- matrix(1:10, ncol=2, byrow=T)
x
#see example matrix

#make the last number 0
x[5,2] <- 0
x

#see which number in the matrix is 0
which(x==0, arr.ind = T)
#arr.ind = T tells which row and column is 0
```


Remove the genes with 0 expression
```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

A common threshold used for calling something differentially expressed is a log2(FoldChange) of greater than 2 or less than -2. Let’s filter the dataset both ways to see how many genes are up or down-regulated.
```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

Determine how many genes are up or down regulated
```{r}
paste("Up:", sum(up.ind))
paste("Down:", sum(down.ind))
```

## Adding annotation data
```{r}
anno <- read.csv("annotables_grch38.csv")
head(anno)
```

Use the merge function
Merge the annotation data with our mycounts data
```{r}
results <- merge(mycounts, anno, by.x="row.names", by.y="ensgene")
head(results)
```

Using Bioconductor's annotation packagse to help with mapping
```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
#download organism annotation package org for homo sapiens "Hs"
#organized as AnnotationDbi database package "db"
#using Entrez Gene IDs "eg" as primary key

columns(org.Hs.eg.db)

mycounts$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(mycounts),
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")

head(mycounts)
```

## DESeq2 analysis
DESeq2 is an R package for analyzing count-based NGS data
```{r}
library(DESeq2)
citation("DESeq2")
```

Create design formula
```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex, 
                              tidy=TRUE)
dds
```

Run DESeq
```{r}
dds <- DESeq(dds)
```

See results
```{r}
res <- results(dds)
res
```

Get summary of results
```{r}
summary(res)
```

Order results table by smallest p value
```{r}
resOrdered <- res[order(res$pvalue),]
```

Adjust p value cut off
```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
```

The more generic way to access the actual subset of the data.frame passing a threshold like this is with the subset() function
```{r}
resSig05 <- subset(as.data.frame(res), padj < 0.05)
nrow(resSig05)
```

Find how many are significant with an adjusted p-val < 0.01
```{r}
resSig01 <- subset(as.data.frame(res), padj < 0.01)
nrow(resSig01)

head(resSig01)
```

Using either the previously generated anno object (annotations from the file annotables_grch38.csv file) or the mapIds() function (from the AnnotationDbi package) add annotation to your res01 results data.frame.
```{r}
columns(org.Hs.eg.db)

resSig01$symbol <- mapIds(org.Hs.eg.db,
                keys=row.names(resSig01),
                column="SYMBOL",
                keytype="ENSEMBL",
                multiVals="first")

resSig01$entrez <- mapIds(org.Hs.eg.db,
                keys=row.names(resSig01),
                column="ENTREZID",
                keytype="ENSEMBL",
                multiVals="first")

resSig01$uniprot <- mapIds(org.Hs.eg.db,
                keys=row.names(resSig01),
                column="UNIPROT",
                keytype="ENSEMBL",
                multiVals="first")
```

Arrange and view results by adjusted p-value
```{r}
ord <- order( resSig01$padj )
head(resSig01[ord,])
```

Write to csv file
```{r}
write.csv(resSig01[ord,], "signif01_results.csv")
```


## Data Visualization
```{r}
#see gene ID for CRISPLD2
i <- grep("CRISPLD2", resSig01$symbol)
resSig01[i,]
```

```{r}
rownames(resSig01[i,])
```

Plot the counts, intgroup means interesting group, variable is the "dex" column
```{r}
plotCounts(dds, gene="ENSG00000103196", intgroup="dex")
```

Return the data
```{r}
d <- plotCounts(dds, gene="ENSG00000103196", intgroup="dex", returnData=TRUE)
head(d)
d
```

Plot with boxplot
```{r}
boxplot(count~dex, data=d)
```

Plot with ggplot2
```{r}
library(ggplot2)
ggplot(d, aes(dex, count)) + 
  geom_boxplot(aes(fill=dex)) + 
  scale_y_log10() + 
  ggtitle("CRISPLD2")
```

To see more plot examples, see 
https://bioboot.github.io/bggn213_S18/class-material/lecture14-BGGN213_W18-lab/




















