---
title: "Untitled"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## PDB Statistics
```{r}
#import data
p <- read.table("Data Export Summary.csv", sep = ",", header = T, row.names = 1)
#set rownames to be experimental method

#calculate percent of each method
percent <- (p$Total / sum(p$Total))*100
names(percent) <- row.names(p)
percent
```


## Using Bio3D
```{r}
library(bio3d)
```

Read in our HIV-Pr structure
```{r}
pdb <- read.pdb("1hsg")
pdb
```

Find attributes of 1hsg
```{r}
attributes(pdb)
```

Access individual attributes or component
```{r}
head(pdb$atom)

#print a subset of $atom data for the first 2 atoms
pdb$atom[1:2, c("eleno", "elety", "x", "y", "z")]

#access individual $atom records
pdb$atom$elety[1:2]

#make plot
plot.bio3d(pdb$atom$b[pdb$calpha], sse=pdb, type = "l", ylab = "B-factor")
```

```{r}
#use str() function to get useful summary of any R object
str(pdb$atom)
```

```{r}
#Print summary of coordinate data in $xyz
pdb$xyz

#examine row and column dimensions
dim(pdb$xyz)

#print coordinates for the first 2 atoms
pdb$xyz[1, atom2xyz((1:2))]
```

Atom selection
```{r}
# Select all C-alpha atoms (return their indices)
ca.inds <- atom.select(pdb, "calpha")
ca.inds

# Print details of the first few selected atoms
head (pdb$atom[ca.inds$atom, ])
```

```{r}
# Print selected xyz coordinates
head(pdb$xyz[, ca.inds$xyz])
```

Various individual atom properties can be used for selection
```{r}
# Select chain A
a.inds <- atom.select(pdb, chain = "A")

# Select C-alphas of chain A
ca.inds <- atom.select(pdb, "calpha", chain = "A")

#Combine multiple selection criteria to return their intersection
cab.inds <- atom.select(pdb, elety=c("CA", "CB"), chain= "A", resno = 10:20)
```

```{r}
ligand.inds <- atom.select(pdb, "ligand")
protein.inds <- atom.select(pdb, "protein")
```


```{r, eval=FALSE}
library(bio3d)
pdb.a <- trim.pdb(pdb, inds=a.inds)
pdb.ca <- trim.pdb(pdb, inds=ca.inds)
pdb.ligand <- trim.pbd(pdb, inds=ligand.inds)
pdb.protein <- trim.pdb(pdb, inds=protein.inds)
```

```{r, eval=FALSE}
write.pdb(pdb.a, file="1hsg_a.pdb")
write.pdb(pdb.ca, file="1hsg_ca.pdb")
write.pbd(pdb.ligand, file="1hsg_ligand.pdb")
write.pdb(pdb.protein, file="1hsg_protein.pdb")
```

## Aligning multiple structures
```{r}
#Download example PDB files
ids <- c("1TND_B","1AGR_A","1TAG_A","1GG2_A","1KJY_A","4G5Q_A")
files <- get.pdb(ids, split = TRUE)
```


```{r}
#extract and align chains we're interested in
pdbs <- pdbaln(files, fit = TRUE)
pdbs
```

```{r}
#access the first 5 rows, and 8 columns
pdbs$ali[1:5, 1:8]

#associated residue numbers
pdbs$resno[1:5, 1:8]
```

Basic structure analysis
```{r}
# Calculate sequence identity
seqidentity(pdbs)

```

```{r}
# Calculate RMSB
rmsd(pdbs)
```


Example application on Adk
```{r}
# Search and retrieve Adenylate kinase structures
aa <- get.seq("1ake_A")
```

```{r}
# Blast or hmmer search
b <- blast.pdb(aa)
```

```{r}
hits <- plot(b)
```

Blast results. Visualize and filter blast results through function plot.blast(). Here
we proceed with only the top scoring hits (black).

```{r}
head(hits$pdb.id)
```

Use function get.pdb() and pdbslit() to fetch and parse the identified
structures. Finally, we use pdbaln() to align the PDB structures.
```{r}
# Fetch PDBs
files <- get.pdb(hits$pdb.id, path = "pdbs", split = TRUE, gzip = TRUE)
```

```{r}
# Align structures
pdbs <- pdbaln(files)
```

```{r}
# Vector containing PDB codes
ids <- basename.pdb(pdbs$id)

# Draw schematic alignment
plot(pdbs, labels=ids)
```
Schematic representation of alignment. Grey regions depict aligned residues, while
white depict gap regions. The red bar at the top depict sequence conservation.


PCA
```{r}
# find invariant core
core <- core.find(pdbs)
#finding rigid core of the proteins

# superimpose all structures to core
pdbs$xyz = pdbfit(pdbs, core)

# Perform PCA
pc.xray <- pca(pdbs)

#Function rmsd() will calculate all pairwise RMSD values of the structural ensemble. This
#facilitates clustering analysis based on the pairwise structural deviation:
# Calculate RMSD
rd <- rmsd(pdbs)

# Structure-based clustering
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd, k=3)
plot(pc.xray, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)
```


```{r}
# Visualize first principal component
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```


View within R with bio3d
```{r, eval=FALSE}
rgl.pdbs <- view(pdbs)
```























